---
title       : Sweave and (perhaps more) knitr course
subtitle    : 
author      : Paola Berchialla
job         : Dept. of Clinical and Biological Science - University of Torino
framework   : io2012     # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
#theme : neon
#transition : horizontal-slide
hitheme     : tomorrow      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}


---
## There was once upon a time Sweave...

- Sweave is a tool that allows to embed R code in (kind of) LaTex documents

- The document will contain both documentation (written in LaTex) and code  (written in R) sections

- The code is evaluated in R

- Figures and tables (your console output) are automatically inserted into the final document

- No need to learn new languages:
  - in the documentation sections, do LaTex
  - in the code sections, do R

---- 
## What is Sweave?
- A set of R functions, written by Friedrich Leisch
(http://www.statistik.lmu.de/~leisch), working under one command in utils package

- Processes R code within a LaTex document

- Returns output from such code 

- Creates plots and automatically creates the LaTex code for their inclusion 

- A LaTex package and style (~\R\R-3.0.2\share\texmf\tex\latex\Sweave.sty)



---
## How to install Sweave
- Assuming LaTex and R are installed, there is no need for installation

- Sweave is distributed with R (since version 1.5.0)

- It is included in the `utils` package (no need to load it)



---
## How does it work?
- Write down the LaTex file with extension .Rnw instead of .tex: example1.Rnw

- The file can contain code chunks, separated from the LaTex pieces

- Within R, run the command Sweave("example1.Rnw")

  -  It executes the code and produces the file example1.tex

  -  Then run LaTex on the example1.tex file to obtain the report


# An easy way is to run Sweave in Rstudio: 
  - open a new file from File, R Sweave
  - just click on compile pdf button after you have edited the .Rnw file

---
## The Noweb syntax
- To separate code and documentation chunks, the **Noweb** syntax is used

- Noweb is a literate programming tool which allows to combine program source code and the corresponding documentation into a single file

- Different segments are called chunks:
  - < < options > >= denotes the start of code chunk,
  - @ denotes the start of a documentation chunk.

- Two kind of operations:
   - weave: typeset documentation together with code
   - tangle: extract code chunks in a .R file

---
## Basic options for code chunks
- *label* is an optional name for the chunk. If it is the first option in the chunk then `label=` can be omitted

- `echo`: if TRUE it echoes the commands, if FALSE it does not (default is TRUE)

- `fig`: if TRUE it includes the plot created in the code (default is FALSE)

---
## A few more options
- `eval`: logical (TRUE). If FALSE, the code chunk is not evaluated, and hence no text or graphical output produced.

- `results`: character string (default is verbatim). 
  - If verbatim, the output of commands is included in the verbatim-like output environment 
  - If tex, the output is taken to be already proper LaTex markup and included as is  
  - If hide then all output is completely suppressed (but the code
executed during the weave)

- `fig`: logical (FALSE), indicating whether the code chunk produces graphical output. Only one figures per code chunk can be processed this way

- `width`: numeric, width of figures in inch

- `height`: numeric, height of figures in inch

- Options can be set globally at the beginning of the file (and changed everywhere else) with \SweaveOpts{option1,option2,...}

---
##  For costly computations

- Check out cacheSweave package

  - Add cache=TRUE to computationally expensive chunks

 	- After the first run, objects from the cached chunks keep stored in a hash map
 
  - Subsequent Sweaves do not evaluate these chunks
  
  -  Cached chunks must be computations only (NO FIGURES)

- In R
```{r, eval=FALSE}
	 library(cacheSweave)
#Set cache directory (default is ".") 
setCacheDir("cache")
# Process document: 
Sweave("mydocument.Rnw", driver = cacheSweaveDriver)
```

---
## Be careful when using cacheSweave
- If data/code changes, you must re-run cached chunks

- Dependencies are not checked: if code in a cached chunk depends on a chunk that has changed, this inconsistency is not detected 

---

## How is `Knitr` better than Sweave?

* _Prettier_ out of the box
	* Code re-formating: use option `tidy`
	* Code highlighting (with Sweave too but you need a _xycolor.sty_ file in your latex distribution)
	* Simple copy-paste enabled
* Better approach to dealing with plots (for example `fig.keep` option for more than 2 plots in 1 chunk)

# knitr main outputs
* `HTML`: via R Markdown files (`Rmd`)
* `PDF`: mostly through `Rnw` files or via `latex` conversion using in R the `pandoc('namefile', format='latex')` command
* Via `pandoc()`, conversion to docx and odt also


---
## R Markdown

The major difference between Rnw (R No web) and Rmd (R Markdown) are the R code chunks

Markdown syntax is very simple and RStudio has a good syntax description

Rnw is still superior for PDF output if you want more control


--- 
## Rmd basics

# Better changing the setting in RStudio from _Sweave_ to _knitr_


  Tools,  Options, Sweave: Weave Rnw files using `knitr`

# Your first Rmd file in Rstudio
1. File, New, R Markdown
2. Click on `MD` (Markdown for quick reference)
3. Edit the title, text, code, add chunks
4. Click on `Knit HTML`

# Rmd R chunks
 In RStudio: Chunks, Insert Chunk
 
 You can add chunk labels after r
	

--- 
## Chunk Options

* Figure
	* `fig.width`, `fig.height`: R control
	* `out.width`, `out.height`: output control
	* `fig.keep`: for more than 2 plots in 1 chunk
	* `fig.cap`: caption in Rnw only	
* Cache
	* `cache`: whether to cache a chunk
	* `dependson`: which other chunks this chunk depends on

more [here] (http://yihui.name/knitr/options)

--- &twocol
## Two column page 

*** left

```{r, echo=F, fig.align="center", fig.height=6, fig.width=4}
set.seed(1234)
x <- rnorm(100)
hist(x, col="light blue", prob=TRUE)
lines(density(x))
```

*** right


```{r, echo=F, fig.align="center", fig.height=6, fig.width=4}
set.seed(1234)
x <- rnorm(100)
hist(x, col="light blue", prob=TRUE)
```


---

## Active development around knitr

# [slidify](http://slidify.org/)
* How this presentation was made

# [knitcitations](https://github.com/cboettig/knitcitations)
* Useful for creating `HTML` citations

# [RStudio](http://www.rstudio.com/)
* Everything works out of the box
* You can use `knit()` instead of `Sweave()` 

# Web Publishing and Blogging
* Easy web publishing tool via [RPubs](http://rpubs.com/)
* `knit2wp()`: knitr to wordpress
* blog with R Markdown and Tumblr: [R Markdown and Tumblr](http://jeffreyhorner.tumblr.com/post/25804518110/blog-with-r-markdown-and-tumblr-part-i)

---

## Commands to make this presentation

```{r eval=FALSE} 

library(slidify)
## Edit index.Rmd file

# run slidify with the following command
slidify("index.Rmd")
```

